---
title: "Survival signature from MOFA model - Factor 2 analysis"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(glmnet)
library(survival)
library(impute)
library(survminer)
library(MOFA2)
library(here)
library(timeROC)
library(survivalROC)
```

### Introduction

The goal is to find the highly discriminatory subset of genes/proteins that provides the maximum separation in survival.

1. **Extract the Leading Edge:** Pull the gene names for the top 15 GSEA pathways we just identified.  This was done in script `gsea_mofa_factor2.rmd`.
2. **Univariate Screening:** Run a quick Cox model on each of those genes individually. Keep only those with $p < 0.05$.
3. **Multivariate Lasso:** Feed those "proven" survival genes into a Lasso-Cox model to find the optimal combination.
4. **C-Index Comparison:** Compare the Concordance Index (C-index) of your new signature against the C-index of Factor 2 alone. If the C-index increases (e.g., from 0.65 to 0.78), you have found a superior discriminator.


### Extract Leading Edge Genes

Leading edge genes from GSEA analysis of MOFA Factor 2 (protein view).
```{r leadingEdgeGenes}
# leading edge genes from top 15 GSEA pathways - see gsea_mofa_factor2.rmd
leading_edge_genes_file <- here("Results", "MOFA_Factor2_Protein_LeadingEdgeGenes_top15PWYs.txt")
leading_edge_genes <- scan(leading_edge_genes_file, what = character())
length(leading_edge_genes)
head(leading_edge_genes)
```

### Survival data and protein expression for leading edge genes
Survival data.
```{r survival_data}
# Load clinical metadata with survival information
clinical_df <- read.csv(here("Data", "Metadata", "metadata_merged_clinical_mol-phenotype_from_paper.csv"), 
                        row.names = 1, 
                        stringsAsFactors = FALSE,
                        na.strings = c("", " ", "NA", "N/A", "na", "null", "NULL", "NaN"))
cat("Before replacement:\n")
print(head(rownames(clinical_df)))

# replace '-' with '.' in case IDs to match sample IDs in MOFA data
rownames(clinical_df) <- gsub("-", ".", rownames(clinical_df))

cat("\nAfter replacement:\n")
print(head(rownames(clinical_df)))
print(dim(clinical_df))

# select relevant columns: case ID, OS_days, OS_event
surv_df <- clinical_df %>%
  select(OS_days, OS_event) %>%
  mutate(OS_days = as.numeric(OS_days), OS_event = as.numeric(OS_event))

head(surv_df)
```

Protein expression data for leading edge genes.
```{r protein_expression}
# Load protein expression data
prot_df <- read.csv(here("Data", "MOFA_input", "proteome_top5k_hvgs_scaled_tumor.csv"), row.names = 1)
print(dim(prot_df))

# subset to leading edge genes
prot_leading_edge_df <- prot_df[rownames(prot_df) %in% leading_edge_genes, ]
print(dim(prot_leading_edge_df))

# check for NAs
na_counts <- rowSums(is.na(prot_leading_edge_df))
# sort and print proteins with any NAs
na_counts <- na_counts[order(na_counts, decreasing = TRUE)]
print(na_counts[na_counts > 0])

# remove proteins with many NAs and no relevant biology (eg housekeeping genes)
proteins_to_remove <- c("ACTN2", "PYGM", "PCSK9", "SH2B1", "TNIP2", "RCL1")
prot_leading_edge_df <- prot_leading_edge_df[!rownames(prot_leading_edge_df) %in% proteins_to_remove, ]
print(dim(prot_leading_edge_df))

# impute remaining NAs, required for glmnet
prot_leading_edge_imputed_df <- as.data.frame(impute.knn(as.matrix(prot_leading_edge_df))$data) 
rownames(prot_leading_edge_imputed_df) <- rownames(prot_leading_edge_df)
dim(prot_leading_edge_imputed_df)
```


### Prepare Data for Lasso-Cox
```{r prepare_data}
# Prepare your Data
# X is a matrix of protein expression for your candidate genes (134 patients x N proteins)
# y is a Surv object: Surv(time, status)
X <- t(prot_leading_edge_imputed_df) 
print(dim(X))

# Surv object: Surv(time, status)
common_ids <- intersect(rownames(X), rownames(surv_df))
X_final <- X[common_ids, ]
surv_final <- surv_df[common_ids, ]
print(dim(X_final))
print(dim(surv_final))

y <- Surv(time = as.numeric(surv_final$OS_days), 
          event = as.numeric(surv_final$OS_event))

# FINAL VERIFICATION
# This MUST be TRUE
all(rownames(X_final) == rownames(surv_final))

# Compare the first 5 values side-by-side
data.frame(
  Input_Days = head(surv_final$OS_days),
  Surv_Object = head(as.matrix(y)[,1])
)
```

### Univariate Screening 
Run a Cox model for each of 355 proteins individually. 
Keep only those with a $p < 0.05$.
```{r univariate_screening}
# Initialize a vector to store p-values
univariate_p <- c()

# Loop through each of your 355 proteins
for(i in seq_len(ncol(X_final))) {
  # Run a simple Cox model: Surv ~ Protein_i
  test_fit <- coxph(y ~ X_final[, i])
  
  # Extract the p-value
  p_val <- summary(test_fit)$coefficients[1, "Pr(>|z|)"]
  univariate_p[colnames(X_final)[i]] <- p_val
}

# 2. Filter the pool to 'Statistically Proven' markers
# We'll use p < 0.05 to be strictly objective
proven_candidates <- names(univariate_p)[univariate_p < 0.05]

# 100 survived
length(proven_candidates)
```

### Clean up missing values
```{r clean_missing}
# Check for NAs in survival data directly
keep_ids <- which(!is.na(surv_final$OS_days) & !is.na(surv_final$OS_event))

# Filter both X and y using the same indices
X_final <- X_final[keep_ids, ]
y_final <- y[keep_ids]
```

### FINAL VERIFICATION
If survival data has any patients with 0 days (which happens in CPTAC if a patient died or was censored the same day as their surgery/biopsy), glmnet will throw this error. 
The Cox model is a ratio-based model that uses logarithms of time; mathematically, $\log(0)$ is undefined, and negative time doesn't exist in biological reality.
```{r zero_day_pts}
# find the "Zero-Day" Patients
# Check for times <= 0
sum(as.matrix(y_final)[,1] <= 0)
# which IDs
which(as.matrix(y_final)[,1] <= 0)

# extract the raw matrix from the Surv object
y_mat <- as.matrix(y_final)

# add a 0.5 day offset to any 0 values - common fix and does not materially affect results
y_mat[y_mat[,1] == 0, 1] <- 0.5

# rebuild the Surv object
y_ultra_final <- Surv(time = y_mat[,1], event = y_mat[,2])

# verify - this should be 0
cat("Times <= 0:", sum(y_ultra_final[,1] <= 0), "\n")
```

### Cross-Validated Lasso-Cox
```{r lasso_cox}
# Subset X to only the 'proven' candidates
X_proven <- X_final[, proven_candidates]
print(dim(X_proven))

set.seed(42) # For reproducibility

# This finds the optimal 'penalty' (lambda)
# use Alpha = 0.9 for stability
# add 'standardize = FALSE' since already scaled
# use 'maxit' to give the engine more time to converge
cv_fit <- cv.glmnet(X_proven, y_ultra_final, 
                    family = "cox", 
                    alpha = 0.9, # <--- ELASTIC NET DEFINED HERE
                    standardize = FALSE, 
                    maxit = 1e6)

# 3. Extract the "Winning" Signature
# 'lambda.min' gives the best prediction, 'lambda.1se' gives the simplest/most robust model
best_lambda <- cv_fit$lambda.min
# Find which index lambda.min corresponds to
which(cv_fit$lambda == cv_fit$lambda.min)
# Extract coefficients at best lambda
coeffs <- coef(cv_fit, s = best_lambda)

# Filter for proteins with non-zero coefficients
optimized_signature <- coeffs[which(coeffs != 0), ]
print(optimized_signature)
```

### Kaplan-Meier Comparison
```{r kaplan_meier}
# Final KM execution
risk_scores <- stats::predict(cv_fit, 
                              newx = X_proven, 
                              s = "lambda.min", 
                              type = "link")

groups <- ifelse(risk_scores > median(risk_scores), "High Risk", "Low Risk")

km_fit <- survfit(y_ultra_final ~ groups)

# The 'survminer' version of the KM plot
ggsurvplot(km_fit, 
           data = data.frame(y_ultra_final),
           pval = TRUE,             # Shows the Log-rank p-value
           conf.int = TRUE,         # Adds the shaded uncertainty area
           risk.table = TRUE,       # Adds the 'Number at Risk' table
           ggtheme = theme_bw(),    # Clean white background
           palette = c("#E41A1C", "#377EB8"),
           title = "Survival by 18-Protein Lasso Signature")
```

### Concordance Index (C-index) Comparison

The **concordance index (C-index)** is a key performance metric in survival analysis and machine learning, measuring an algorithm's ability to correctly rank patient risk scores or survival times.

It evaluates pairs of subjects, determining if the subject with the higher predicted risk actually experienced the event first. In survival analysis, a **concordant pair** consists of two subjects where the model's predicted risk correctly orders their actual survival times. If Subject A has a higher predicted risk (shorter predicted survival) and actually experiences the event first (dies earlier) than Subject B, the pair is concordant. It indicates agreement between model predictions and observed outcomes. 

#### Concordance Index Score Interpretation:
*   **0.5**: No better than random chance.
*   **0.7–0.8**: Generally considered acceptable or good.
*   **0.8–0.9**: Strong discrimination.
*   **> 0.9**: Excellent, sometimes suggesting overfitting. 













