---
title: "Validation of Signatures in APGI Cohort"
subtitle: "Two Signatures: Tong ≥80% and Both Cohorts ≥80%"
author: "CPTAC-PDA Analysis"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(survival)
library(survminer)
library(here)
```

### Introduction

Validation of two Lasso-Cox prognostic signatures (derived from CPTAC-PDAC) using the independent APGI/ProCan cohort:

1. **Tong ≥80% Signature** — proteins detected in ≥80% of Tong tumors (discovery C-index: 0.707)
2. **Both Cohorts ≥80% Signature** — proteins detected in ≥80% of both Tong & APGI (discovery C-index: 0.700)

**Methodology:**
1. Load APGI protein matrix and map to gene symbols.
2. Z-score normalize the APGI data (gene-wise) to match CPTAC processing.
3. Apply Lasso coefficients from each signature to calculate Risk Scores.
4. Perform survival analysis for each.

---

## 1. Load Data

```{r load_data}
# APGI raw protein matrix (uniprot IDs as columns)
apgi_raw <- read.csv(here("Data", "APGI-PDA", "APGI_protein_matrix_and_metadata.csv"), 
                     row.names = 1, check.names = FALSE)

cat("APGI raw:", nrow(apgi_raw), "samples x", ncol(apgi_raw), "columns\n")

# Separate protein data from clinical metadata
# Last protein column is "Q9Y6Y8", first clinical column is "Prep_LIMS_ID"
last_prot <- "Q9Y6Y8"
prot_cols <- 1:which(names(apgi_raw) == last_prot)

apgi_prot_uniprot <- apgi_raw[, prot_cols]
apgi_clin <- apgi_raw[, -prot_cols]

cat("Protein columns:", ncol(apgi_prot_uniprot), "\n")
cat("Clinical columns:", ncol(apgi_clin), "\n")

# Map uniprot IDs to gene symbols
apgi_gene_map <- read.csv(here("Data", "APGI-PDA", "APGI_protein_and_gene_names.csv"))
uniprot_to_gene <- setNames(apgi_gene_map$Genes, apgi_gene_map$UniprotID)
mapped <- uniprot_to_gene[names(apgi_prot_uniprot)]
mapped <- mapped[!is.na(mapped)]

apgi_prot <- apgi_prot_uniprot[, names(mapped)]
colnames(apgi_prot) <- as.character(mapped)
cat("Mapped to gene symbols:", ncol(apgi_prot), "proteins\n")

# Signature files
sig_files <- c(
  "MOFA_Factor2_LassoSignature_Tong_ge80pct_Tumors.csv",
  "MOFA_Factor2_LassoSignature_Both_Cohorts_ge80pct.csv"
)
sig_labels <- c("Tong >=80% Signature", "Both Cohorts >=80% Signature")
```

### Data Preprocessing

```{r preprocessing}
# APGI: gene-wise Z-score normalization
apgi_z <- as.data.frame(scale(apgi_prot))

# Clinical: survival data
apgi_clin <- apgi_clin %>%
  mutate(
    OS_days = as.numeric(`Length of Follow up (days)`),
    OS_event = ifelse(grepl("Deceased", Status), 1, 0)
  )
```

---

## 2. Validation Function

```{r validation_function}
validate_signature <- function(sig_file, sig_label) {
  cat("\n========================================\n")
  cat("Validating:", sig_label, "\n")
  cat("========================================\n")
  
  # Load signature
  sig_df <- read.csv(here("Results", sig_file))
  sig_proteins <- sig_df$Protein
  cat("Signature proteins:", length(sig_proteins), "\n")
  
  # Intersect with APGI
  common <- intersect(colnames(apgi_z), sig_proteins)
  missing <- setdiff(sig_proteins, colnames(apgi_z))
  cat("Found in APGI:", length(common), "/", length(sig_proteins), "\n")
  if (length(missing) > 0) cat("Missing:", paste(missing, collapse = ", "), "\n")
  
  if (length(common) < 3) {
    cat("Too few proteins, skipping\n")
    return(NULL)
  }
  
  # Only use coefficients for proteins present in APGI
  sig_subset <- sig_df[sig_df$Protein %in% common, ]
  
  # Calculate Risk Score
  risk_scores <- as.matrix(apgi_z[, common]) %*% sig_subset$Coefficient
  risk_scores <- as.numeric(risk_scores[, 1])
  names(risk_scores) <- rownames(apgi_z)
  
  # Handle NAs: if any protein has NAs, the risk score is NA for that sample
  na_idx <- is.na(risk_scores)
  cat("Samples with NA risk scores:", sum(na_idx), "/", length(risk_scores), "\n")
  
  # Build analysis dataframe
  analysis_df <- data.frame(
    SampleID = rownames(apgi_z),
    RiskScore = risk_scores,
    OS_days = apgi_clin$OS_days,
    OS_event = apgi_clin$OS_event
  )
  analysis_df_clean <- analysis_df[!is.na(analysis_df$RiskScore), ]
  
  # Median split for KM
  median_rs <- median(analysis_df_clean$RiskScore)
  analysis_df_clean$RiskGroup <- ifelse(
    analysis_df_clean$RiskScore > median_rs, "High Risk", "Low Risk"
  )
  
  cat("Samples after NA removal:", nrow(analysis_df_clean), "\n")
  cat("Median RiskScore:", round(median_rs, 3), "\n")
  cat("High Risk:", sum(analysis_df_clean$RiskGroup == "High Risk"), "\n")
  cat("Low Risk:", sum(analysis_df_clean$RiskGroup == "Low Risk"), "\n")
  
  # --- Cox PH ---
  y <- Surv(time = analysis_df_clean$OS_days, event = analysis_df_clean$OS_event)
  cox_fit <- coxph(y ~ RiskScore, data = analysis_df_clean)
  cox_sum <- summary(cox_fit)
  
  hr <- cox_sum$conf.int[1]
  hr_lower <- cox_sum$conf.int[3]
  hr_upper <- cox_sum$conf.int[4]
  hr_p <- cox_sum$coefficients[5]
  
  cat(sprintf("HR: %.2f (%.2f-%.2f), p = %.1e\n", hr, hr_lower, hr_upper, hr_p))
  
  # --- C-index ---
  c_index <- concordance(y ~ analysis_df_clean$RiskScore)$concordance
  c_index <- ifelse(c_index < 0.5, 1 - c_index, c_index)
  cat("C-index:", round(c_index, 3), "\n")
  
  # --- KM ---
  km_fit <- survfit(y ~ RiskGroup, data = analysis_df_clean)
  
  # --- Log-rank ---
  lr <- survdiff(y ~ RiskGroup, data = analysis_df_clean)
  logrank_p <- 1 - pchisq(lr$chisq, df = 1)
  
  # --- Results ---
  results <- data.frame(
    Signature = sig_label,
    N_Proteins_Original = length(sig_proteins),
    N_Proteins_Validated = length(common),
    Missing_Proteins = paste(missing, collapse = "; "),
    N_Patients = nrow(analysis_df_clean),
    HR = round(hr, 3),
    HR_Lower = round(hr_lower, 3),
    HR_Upper = round(hr_upper, 3),
    HR_pval = format(hr_p, digits = 4),
    C_Index = round(c_index, 3),
    LogRank_pval = format(logrank_p, digits = 4)
  )
  
  return(list(
    label = sig_label,
    results = results,
    km_fit = km_fit,
    analysis_df = analysis_df_clean,
    y = y,
    cox_fit = cox_fit,
    hr = hr, hr_lower = hr_lower, hr_upper = hr_upper, hr_p = hr_p,
    c_index = c_index, logrank_p = logrank_p
  ))
}
```

---

## 3. Run Validation for Both Signatures

```{r run_validation}
val_results <- list()
for (i in seq_along(sig_files)) {
  r <- validate_signature(sig_files[i], sig_labels[i])
  if (!is.null(r)) val_results[[length(val_results) + 1]] <- r
}
```

---

## 4. Summary Table

```{r summary_table}
metrics_df <- do.call(rbind, lapply(val_results, function(r) r$results))
knitr::kable(metrics_df, caption = "APGI Validation Results", digits = 3)
```

---

## 5. Kaplan-Meier Curves

```{r km_curves, fig.width=8, fig.height=6}
for (res in val_results) {
  df <- res$analysis_df
  km_fit <- survfit(Surv(OS_days, OS_event) ~ RiskGroup, data = df)
  p <- ggsurvplot(km_fit, data = df, pval = TRUE, risk.table = TRUE,
                  conf.int = TRUE,
                  title = paste("APGI Validation:", res$label),
                  xlab = "Time (days)", ylab = "Overall Survival",
                  palette = c("#E41A1C", "#377EB8"),
                  legend.title = "Risk Group",
                  legend.labs = c("High Risk", "Low Risk"))
  print(p)
}
```

---

## 6. Export Results

```{r export, include=FALSE}
write.csv(metrics_df, here("Results", "APGI_Validation_MetricsSummary.csv"), row.names = FALSE)
cat("Exported to Results/APGI_Validation_MetricsSummary.csv\n")
```
