---
title: "Prepare Tong Cohort for 18-Protein Signature Validation"
subtitle: "Data preparation of Tong et al. (J Hematol Oncol, 2022) proteomics"
author: "CPTAC-PDA Analysis"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(readxl)
```

## Overview

Prepares the Tong et al. (2022) proteomics cohort for external validation of the CPTAC-PDA 18-protein Lasso-Cox signature.

**Data sources:**
1. CPTAC Signature: 18 proteins (Gene Symbols) with Lasso-Cox coefficients
2. Tong Table S3A: 7,055 protein groups in 226 tumors + 220 NATs (at least 1/6 of samples)

**Workflow:** Load Tong data -> segregate tumors -> Z-score normalize -> extract signature -> save

---

## 1. Load CPTAC Signature

```{r load_signature}
signature_df <- data.frame(
  Protein = c("HMGCS2", "ADH1C", "F3", "TNC", "MYH11", "TK1", "PRKCB", 
              "KPNA2", "C1QC", "GNAO1", "CBR1", "IFITM3", "CEBPB", 
              "HLA-E", "PCNA", "SHMT2", "MRTO4", "IDH1"),
  Coefficient = c(-0.1929, -0.01069, 0.1508, 0.06905, -0.06486, 0.03575,
                  -0.06545, 0.02124, 0.1214, -0.03794, -0.2453, 0.08542,
                  0.06982, 0.07697, 0.07518, 0.02068, 0.1733, -0.07292))
print(signature_df)
cat("\nNote: Gene Symbols already. No UniProt mapping needed.\n")
```

---

## 2. Load Tong Data

```{r load_tong}
tong_raw <- read_excel("13045_2022_1384_MOESM25_ESM.xlsx", sheet = "Table S3A ")
cat("Dim:", dim(tong_raw), "\n")
cat("First col:", names(tong_raw)[1], "\n")
```

Tumor columns end with T, normal with N.

```{r segregate}
t_cols <- grep("T$", names(tong_raw), value = TRUE)
n_cols <- grep("N$", names(tong_raw), value = TRUE)
cat("Tumor samples:", length(t_cols), "\n")
cat("Normal samples:", length(n_cols), "\n")

# Build tumor matrix
symbols <- tong_raw[[1]]
tumor_df <- tong_raw[, t_cols]
tumor_df <- as.data.frame(lapply(tumor_df, as.numeric))
rownames(tumor_df) <- symbols
cat("Tumor matrix:", dim(tumor_df), "\n")
```

---

## 3. Z-score Normalize (CPTAC-style)

CPTAC: filter <=20% NAs, then row-wise Z-score.

```{r normalize}
# Filter
na_pct <- rowSums(is.na(tumor_df)) / ncol(tumor_df) * 100
keep <- na_pct <= 20
tumor_filt <- tumor_df[keep, ]
cat("After 20% NA filter:", nrow(tumor_filt), "proteins\n")

# Vectorized Z-score
tm <- as.matrix(tumor_filt)
means <- rowMeans(tm, na.rm = TRUE)
sds <- apply(tm, 1, sd, na.rm = TRUE)
tumor_scaled <- sweep(sweep(tm, 1, means, "-"), 1, sds, "/")

cat("Verification - row 1 mean:", round(mean(tumor_scaled[1,], na.rm=TRUE), 6), "\n")
cat("Verification - row 1 sd:", round(sd(tumor_scaled[1,], na.rm=TRUE), 6), "\n")
```

---

## 4. Extract Signature

```{r extract}
found <- intersect(signature_df$Protein, rownames(tumor_scaled))
missing <- setdiff(signature_df$Protein, rownames(tumor_scaled))
cat("Found:", length(found), "/ 18\n")
cat("Missing:", paste(missing, collapse=", "), "\n")

sig_mat <- tumor_scaled[found, , drop = FALSE]
```

```{r impute}
# Check and impute NAs
sig_na <- round(rowSums(is.na(sig_mat)) / ncol(sig_mat) * 100, 1)
print(data.frame(Protein=names(sig_na), MissingPct=sig_na))

sig_imputed <- sig_mat
for (i in seq_len(nrow(sig_imputed))) {
  idx <- which(is.na(sig_imputed[i,]))
  if (length(idx) > 0) {
    sig_imputed[i, idx] <- median(as.numeric(sig_imputed[i,]), na.rm=TRUE)
  }
}
cat("Remaining NAs:", any(is.na(sig_imputed)), "\n")
```

---

## 5. Calculate Risk Scores & Save

```{r save}
# Transpose
validation <- as.data.frame(t(sig_imputed))
colnames(validation) <- found

# Align coefficients
coeffs <- signature_df$Coefficient[match(found, signature_df$Protein)]
cat("Alignment OK:", all(colnames(validation) == found), "\n")

# Risk scores
validation$RiskScore <- as.matrix(validation) %*% coeffs
validation$SampleID <- rownames(validation)

cat("RiskScore summary:\n")
print(summary(validation$RiskScore))

write.csv(validation, "tong_validation_data_prepared.csv", row.names=FALSE)
write.csv(as.data.frame(t(sig_imputed)), "tong_signature_proteins_scaled.csv", row.names=TRUE)
cat("\nSaved.\n")
```

---

## Summary

```{r summary}
cat("Tong validation data prepared.\n")
cat("Samples:", nrow(validation), "tumors\n")
cat("Signature proteins:", length(found), "/ 18\n")
cat("Missing:", if(length(missing)>0) paste(missing,collapse=", ") else "none", "\n")
```
