---
title: "Prepare Tong Cohort for Signature Validation"
subtitle: "Two Signatures: Tong ≥80% and Both Cohorts ≥80%"
author: "CPTAC-PDA Analysis"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(readxl)
library(here)
```

## Overview

Prepares the Tong et al. (2022) proteomics cohort for external validation of two Lasso-Cox signatures:

1. **Tong ≥80% Signature** (proteins detected in ≥80% of Tong tumors)
2. **Both Cohorts ≥80% Signature** (proteins detected in ≥80% of both Tong & APGI)

**Workflow:** Load Tong data → segregate tumors → Z-score normalize → extract each signature → save risk scores

---

## 1. Load Signatures

```{r load_signatures}
sig_files <- c(
  "MOFA_Factor2_LassoSignature_Tong_ge80pct_Tumors.csv",
  "MOFA_Factor2_LassoSignature_Both_Cohorts_ge80pct.csv"
)
sig_labels <- c("Tong_ge80pct", "Both_ge80pct")

sig_list <- list()
for (i in seq_along(sig_files)) {
  df <- read.csv(here("Results", sig_files[i]))
  sig_list[[sig_labels[i]]] <- df
  cat(sig_labels[i], ":", nrow(df), "proteins\n")
}
```

---

## 2. Load Tong Data

```{r load_tong}
tong_raw <- read_excel(here("Data", "Tong-PDA", "13045_2022_1384_MOESM25_ESM.xlsx"), 
                       sheet = "Table S3A ")
cat("Tong raw dim:", dim(tong_raw), "\n")

# Tumor columns end with T
t_cols <- grep("T$", names(tong_raw), value = TRUE)
cat("Tumor samples:", length(t_cols), "\n")

# Build tumor matrix
symbols <- as.character(tong_raw[[1]])
tumor_df <- tong_raw[, t_cols]
tumor_df <- as.data.frame(lapply(tumor_df, as.numeric))
rownames(tumor_df) <- symbols
cat("Tumor matrix:", dim(tumor_df), "\n")
```

---

## 3. Z-score Normalize

```{r normalize}
# Filter: keep proteins with <=20% NAs
na_pct <- rowSums(is.na(tumor_df)) / ncol(tumor_df) * 100
keep <- na_pct <= 20
tumor_filt <- tumor_df[keep, ]
cat("After 20% NA filter:", nrow(tumor_filt), "proteins\n")

# Row-wise Z-score
tm <- as.matrix(tumor_filt)
means <- rowMeans(tm, na.rm = TRUE)
sds <- apply(tm, 1, sd, na.rm = TRUE)
tumor_scaled <- sweep(sweep(tm, 1, means, "-"), 1, sds, "/")
```

---

## 4. Extract Each Signature & Compute Risk Scores

```{r extract_save}
for (sig_label in sig_labels) {
  cat("\n========", sig_label, "========\n")
  
  sig_df <- sig_list[[sig_label]]
  sig_proteins <- sig_df$Protein
  
  # Find which signature proteins are available in Tong
  found <- intersect(sig_proteins, rownames(tumor_scaled))
  missing <- setdiff(sig_proteins, rownames(tumor_scaled))
  cat("Found:", length(found), "/", length(sig_proteins), "\n")
  if (length(missing) > 0) cat("Missing:", paste(missing, collapse = ", "), "\n")
  
  if (length(found) < 3) {
    cat("Too few proteins, skipping\n")
    next
  }
  
  # Extract and impute
  sig_mat <- tumor_scaled[found, , drop = FALSE]
  sig_imputed <- sig_mat
  for (i in seq_len(nrow(sig_imputed))) {
    idx <- which(is.na(sig_imputed[i, ]))
    if (length(idx) > 0) {
      sig_imputed[i, idx] <- median(as.numeric(sig_imputed[i, ]), na.rm = TRUE)
    }
  }
  
  # Align coefficients to match available proteins
  coeffs <- sig_df$Coefficient[match(found, sig_df$Protein)]
  
  # Risk scores
  validation <- as.data.frame(t(sig_imputed))
  colnames(validation) <- found
  validation$RiskScore <- as.matrix(validation) %*% coeffs
  validation$SampleID <- rownames(validation)
  
  cat("RiskScore summary:\n")
  print(summary(validation$RiskScore))
  
  # Save
  outfile <- paste0("tong_validation_", sig_label, ".csv")
  write.csv(validation, here("Data", "Tong-PDA", outfile), row.names = FALSE)
  cat("Saved to Data/Tong-PDA/", outfile, "\n", sep = "")
}
```

---

## Summary

```{r summary}
cat("Tong validation data prepared for:\n")
for (sig_label in sig_labels) {
  outfile <- paste0("tong_validation_", sig_label, ".csv")
  cat("  -", sig_label, ":", outfile, "\n")
}
```
