---
title: "External Validation of 18-Protein Lasso-Cox Signature on Tong Cohort"
subtitle: "Survival Analysis: Kaplan-Meier, C-index, Time-Dependent ROC"
author: "CPTAC-PDA Analysis"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(survival)
library(survminer)
library(readxl)
```

## Overview

Validates the 18-protein Lasso-Cox signature from CPTAC-PDA on the independent Tong et al. (2022) cohort. The signature was derived from 140 CPTAC-PDA patients and is tested here on 226 Tong PDAC patients.

**Key analyses:**
1. Kaplan-Meier survival curves (median split)
2. Concordance Index (C-index)
3. Time-Dependent ROC at 1 and 2 years

---

## 1. Load Data

```{r load}
# Validation data (prepared in 01_prepare_tong_validation_data.Rmd)
val <- read.csv(here("Data", "Tong-PDA", "tong_validation_data_prepared.csv"))
cat("Validation data:", nrow(val), "samples\n")
cat("Proteins:", sum(!names(val) %in% c("RiskScore", "SampleID")), "\n")

# Survival data
surv_raw <- read_excel(here("Data", "Tong-PDA", "13045_2022_1384_MOESM23_ESM.xlsx"), sheet = "Table S1B")
surv <- data.frame(
  SampleID = as.character(surv_raw[[1]]),
  OS_days = as.numeric(as.character(surv_raw[["Overall Survival (Months)"]])),
  OS_event = as.numeric(as.character(surv_raw[["Status of OS ( 1: death; 0: survival or loss to follow up )"]]))
)
cat("Survival data:", nrow(surv), "samples\n")

# Merge: strip 'T' suffix from validation IDs to match survival IDs
val$PatientID <- gsub("T$", "", val$SampleID)
merged <- inner_join(val, surv, by = c("PatientID" = "SampleID"))
cat("\nMerged dataset:", nrow(merged), "patients\n")
```

---

## 2. Kaplan-Meier Analysis

Median risk score split to define High and Low risk groups, matching CPTAC-PDA validation.

```{r km}
# Median split
median_risk <- median(merged$RiskScore)
merged$RiskGroup <- ifelse(merged$RiskScore > median_risk, "High Risk", "Low Risk")
cat("Median risk score:", round(median_risk, 4), "\n")
cat("High Risk:", sum(merged$RiskGroup == "High Risk"), "\n")
cat("Low Risk:", sum(merged$RiskGroup == "Low Risk"), "\n")

# Convert OS from months to days (for consistency with CPTAC analysis)
merged$OS_time <- merged$OS_days * 30.4375  # average days per month

# Fit KM
km_fit <- survfit(Surv(OS_time, OS_event) ~ RiskGroup, data = merged)

# Plot
ggsurvplot(km_fit, 
           data = merged,
           pval = TRUE,
           conf.int = TRUE,
           risk.table = TRUE,
           ggtheme = theme_bw(),
           palette = c("#E41A1C", "#377EB8"),
           title = "Tong Cohort: 11-Protein Signature Survival",
           xlab = "Time (days)",
           ylab = "Overall Survival Probability")
```

```{r median_summary}
# Median survival by group
km_summary <- surv_median(km_fit)
print(km_summary)

# Log-rank test
lr <- survdiff(Surv(OS_time, OS_event) ~ RiskGroup, data = merged)
p_val <- 1 - pchisq(lr$chisq, df = 1)
cat("\nLog-rank p-value:", format(p_val, digits = 4), "\n")

# Cox PH model for Hazard Ratio
cox_model <- coxph(Surv(OS_time, OS_event) ~ RiskGroup, data = merged)
cox_summary <- summary(cox_model)
hr <- cox_summary$conf.int[1]
hr_lower <- cox_summary$conf.int[3]
hr_upper <- cox_summary$conf.int[4]
hr_p <- cox_summary$waldtest[3]

cat("\nHazard Ratio (High vs Low Risk):", round(hr, 4), "\n")
cat("95% CI:", round(hr_lower, 4), "-", round(hr_upper, 4), "\n")
cat("p-value:", format(hr_p, digits = 4), "\n")

# Continuous RiskScore HR
cox_cont <- coxph(Surv(OS_time, OS_event) ~ RiskScore, data = merged)
cat("\nHR per unit RiskScore:", round(exp(coef(cox_cont)), 4), 
    "(p =", format(summary(cox_cont)$waldtest[3], digits = 4), ")\n")
```

---

## 3. Concordance Index (C-index)

```{r cindex}
# C-index for the signature risk score
c_index <- concordance(Surv(OS_time, OS_event) ~ RiskScore, data = merged)
cat("Raw C-index:", round(c_index$concordance, 4), "\n")
cat("Adjusted C-index (flip if <0.5):", 
    round(ifelse(c_index$concordance < 0.5, 1 - c_index$concordance, c_index$concordance), 4), "\n")

# Compare with CPTAC cohort results
cat("\n--- Comparison ---\n")
cat("CPTAC (18-protein): True C-index = 0.7359\n")
cat("Tong (11-protein):  True C-index =", 
    round(ifelse(c_index$concordance < 0.5, 1 - c_index$concordance, c_index$concordance), 4), "\n")
```

---

## 4. Time-Dependent ROC

```{r roc, warning=FALSE}
# Check if survivalROC is available
if (!require("survivalROC", quietly = TRUE)) {
  install.packages("survivalROC", repos = "https://cran.r-project.org")
  library(survivalROC)
}

# 1-year and 2-year ROC (using months for interpretability)
res_1yr <- survivalROC(
  Stime = merged$OS_days,
  status = merged$OS_event,
  marker = merged$RiskScore,
  predict.time = 12,
  method = "KM"
)

res_2yr <- survivalROC(
  Stime = merged$OS_days,
  status = merged$OS_event,
  marker = merged$RiskScore,
  predict.time = 24,
  method = "KM"
)

# Plot
plot(res_1yr$FP, res_1yr$TP, type = "l", col = "#E41A1C", lwd = 3,
     xlim = c(0, 1), ylim = c(0, 1),
     xlab = "1 - Specificity", ylab = "Sensitivity",
     main = "Tong Cohort: Time-Dependent ROC")
lines(res_2yr$FP, res_2yr$TP, col = "#377EB8", lwd = 3)
abline(0, 1, lty = 2, col = "darkgrey")

legend("bottomright",
       c(paste0("1-Year (AUC = ", round(res_1yr$AUC, 2), ")"),
         paste0("2-Year (AUC = ", round(res_2yr$AUC, 2), ")")),
       col = c("#E41A1C", "#377EB8"), lty = 1, lwd = 3)
```

---

## Summary

```{r summary}
cat("========================================\n")
cat("TONG COHORT VALIDATION - RESULTS\n")
cat("========================================\n\n")
cat("Sample size:", nrow(merged), "\n")
cat("Signature proteins available:", 
    sum(!names(val) %in% c("RiskScore", "SampleID", "PatientID")), "/ 18\n\n")

cat("Kaplan-Meier:\n")
cat("  Median survival (High Risk):", round(km_summary$median[1], 1), "days\n")
cat("  Median survival (Low Risk):", round(km_summary$median[2], 1), "days\n")
cat("  Log-rank p:", format(p_val, digits = 4), "\n\n")

c_adj <- ifelse(c_index$concordance < 0.5, 1 - c_index$concordance, c_index$concordance)
cat("C-index (adjusted):", round(c_adj, 4), "\n\n")

cat("Time-Dependent AUC:\n")
cat("  1-year:", round(res_1yr$AUC, 3), "\n")
cat("  2-year:", round(res_2yr$AUC, 3), "\n")
cat("========================================\n")
```

---

## Session Info

```{r}
sessionInfo()
```
